---
title: "Data Analysis for the Thesis"
author: "Anita Di Gennaro"
date: "2026-04-27"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# PHASE 0

This initial phase 0 deals with the data cleaning and preparation of the
real-data backbone of the thesis from two ISTAT sources: SES income data
and RACLI regional data from 2022. The output is three tidy datasets:
*'ses_data', 'racli_clean'*, and *'regional_codes',* that together
describe the structure of Italian wages along the demographic, sectoral,
and geographic macro and micro areas, which will be used as the
experiment's backbone.

In the initial sections, the SES (Structure of Earnings Survey) data was
elaborated and cleaned: a single *parse_ses_table()* function, applied
across the 15 relevant sheets via a metadata table, was used to extract
and label every cell into one long tidy data frame. SES describes wages
by sector, firm size, control type, contract, hours regime, education,
occupation (ISCO), and seniority, but resolves geography only to five
NUTS-1 geographical macro-areas.

Later on, the RACLI regional wage file was filtered to 2022, stripped
from 26 columns to 8 by removing constant and all-NA fields, and recoded
into the same sex and geographic conventions as SES so that the two can
follow a common structure. RACLI describes wages by sex, year, and
quantile (mean, p10, median, p90) across all 20 NUTS-2 regions and 107
NUTS-3 provinces, but does not contain any occupational, sectoral, or
demographic details.

A small bridge table (*regional_codes*) maps each NUTS-2 region to its
NUTS-1 macro-area, accommodating the fact that RACLI still uses the
pre-2021 NUTS coding (ITD, ITE) while the modern standard uses ITH and
ITI. Neither data set alone has the joint distribution of demographics ×
sector × NUTS-2 region; which could potentially be the gap filled with
the LLMs (?).

```{r include=FALSE}
library(tidyr)
library(readxl)
library(dplyr)
library(tidyverse)
library(stringr)
library(readr)
library(ggplot2)
library(DataExplorer)
```

# SES Wage Structure

This is a series of aggregated tables. In order to understand how to
navigate them, I will start by uploading the excel sheets and looking at
them to then manually work on them based on what I see/need.

```{r}
# this will read the index sheet
# we can see that the first two are indexes - so skip
ses_2022 <- read_excel("data/Tavole_Struttura-delle-retribuzioni_2022.xlsx",
                       skip = 2) # skip blank rows on top


all_sheets <- excel_sheets("data/Tavole_Struttura-delle-retribuzioni_2022.xlsx") # list of all sheets
print(ses_2022)
```

There are 16 tables. Let's understand the layout of one table to see how
to filter that one and move around it.

# Test on Table 1 (tav 1)

```{r}
tav1_raw <- read_excel("data/Tavole_Struttura-delle-retribuzioni_2022.xlsx",
                       sheet = "Tav.1",
                       col_names = FALSE,
                       col_types = "text")   # force everything to text and avoids type-guessing issues



# rename
tav1_col_names <- c( "row_label", "F_educ_low", "F_educ_med", "F_educ_high", "F_total", "sep1",
                     "M_educ_low", "M_educ_med", "M_educ_high", "M_total", "sep2",
                     "T_educ_low", "T_educ_med", "T_educ_high", "T_total")

tav1_raw <- read_excel("data/Tavole_Struttura-delle-retribuzioni_2022.xlsx",
                       sheet = "Tav.1",
                       col_names = tav1_col_names,
                       col_types = "text",
                       skip = 4)
```

## Section types and filter to data rows

The four section markers act as "headers" for the rows below them. The
main idea is to identify them, use fill() to propagate the label
downward, then remove the marker rows themselves and the footnotes.

```{r}
sections <- c("ATTIVITÀ ECONOMICA","ATTIVITÀ ECONOMICA ",   # ISTAT has a trailing space in some tables so ATTIVITA' ECONOMICA is repeated with a space afterwards too
              "RIPARTIZIONE GEOGRAFICA",
              "CLASSE DIMENSIONALE", 
              "TIPO DI CONTROLLO ECONOMICO")

tav1_clean <- tav1_raw |>
  mutate(
    # when col 1 is NA and col 2 matches a known marker, record the section name
    strat_type = if_else(
      is.na(row_label) & str_trim(F_educ_low) %in% sections,
      str_trim(F_educ_low),
      NA_character_)) |>
  # add section label downward to all rows belonging to that section
  fill(strat_type, .direction = "down") |>
  # drop the section marker rows (row_label is NA) and footnote rows
  filter(!is.na(row_label),
         !str_detect(row_label, "^legenda|^Fonte"),
         !str_detect(row_label, "^None$"))  # in case of leftover "None" strings
  
 
# shorten the section labels to something clean
tav1_clean <- tav1_clean |>
  mutate(strat_type = recode(str_trim(strat_type),
    "ATTIVITÀ ECONOMICA" = "sector",
    "ATTIVITÀ ECONOMICA "  = "sector",
    "RIPARTIZIONE GEOGRAFICA" = "geography",
    "CLASSE DIMENSIONALE" = "firm_size",
    "TIPO DI CONTROLLO ECONOMICO" = "control_type"))

# rename the levels of education without specifying they refer to education, as I will address it as the table's breakdown category
tav1_clean <- tav1_clean |>
  rename(
    F_low  = F_educ_low,
    F_med  = F_educ_med,
    F_high = F_educ_high,
    M_low  = M_educ_low,
    M_med  = M_educ_med,
    M_high = M_educ_high,
    T_low  = T_educ_low,
    T_med  = T_educ_med,
    T_high = T_educ_high)

tav1_clean
```

## Convert values and drop separator columns

Drop NAs, convert to numeric what is character and turn into NAs "..",
which ISTAT uses to indicate suppressed data (cell too small to
publish). Lastly, the table will be pivoted and the names recoded to be
clearer.

```{r}
tav1_clean <- tav1_clean |>
  select(-c(sep1, sep2)) |>
  mutate(across(.cols = c(F_low:T_total),
                .fns  = ~ as.numeric(if_else(str_trim(.) == "..", NA_character_, .))
                ))

# pivot and mutate
tav1_final <- tav1_clean |>
  pivot_longer(
    cols      = c(F_low:T_total),
    names_to  = c("sex", "breakdown_cat"),
    names_sep = "_",          # split on first underscore 
    values_to = "value") |>
  mutate(sex = recode(sex,
                      "F" = "female",
                      "M" = "male",
                      "T" = "total"),
    breakdown_var = "education",
    outcome_var = "hourly_wage") 
# label what the breakdown variable is and the outcome variable 
  

tav1_final
```

Now the table 1 finalized has the labelled sector as an indicator, the
stratification type, the sex, the reference category (according to which
this table is broken down and categorized - therefore the
breakdown_cat), and the outcome variable, what the values correspond to.

This is a test performed to Table 1 to see if this is a format that
could then be generalized to the rest of the tables, when a complete
data set will be created.

In order for the complete data set to be created, I need to label the
breakdown variables that are used across the tables. SES classifies the
tables according to meta-level variables: education (like tav 1),
contract type, occupation (ISCO), hours regime, seniority brackets.

```{r recoding}
# education (ISCED)
meta_education_level <- list(raw_names = c("row_label","F_low", "F_med", "F_high", 
                                           "F_total", "sep1",
                                           "M_low", "M_med", "M_high", "M_total", "sep2",
                                           "T_low", "T_med", "T_high", "T_total"),
  breakdown_var = "education",
  cat_labels = c(low = "≤ lower secondary",
                    med = "upper secondary",
                    high = "tertiary",
                    total = "total"))
 
# contract type (Tav.2, 7, 12)
meta_contract_type <- list(raw_names = c(
  "row_label",
  "F_permanent", "F_temporary", "F_total", "sep1",
  "M_permanent", "M_temporary", "M_total", "sep2",
  "T_permanent", "T_temporary", "T_total"),
  breakdown_var = "contract_type",
  cat_labels = c(permanent = "permanent", temporary = "temporary", total = "total"))
 
# hours regime (Tav.3, 8, 13)
meta_hours <- list(raw_names = c(
  "row_label",
  "F_parttime", "F_fulltime", "F_total", "sep1",
   "M_parttime", "M_fulltime", "M_total", "sep2",
  "T_parttime", "T_fulltime", "T_total"),
  breakdown_var = "hours_regime",
  cat_labels = c(parttime = "part-time", fulltime = "full-time", total = "total"))
 
# occupation (ISCO) (Tav.4, 9, 15)
# 10 ISCO major groups + total for each sex
# (Forze Armate is almost always ".." but it was included it as its own thing)
meta_ISCO <- list(raw_names = c(
    "row_label",
    "F_armed", "F_manager", "F_intellectual", "F_technical",
    "F_clerical", "F_sales", "F_agriculture", "F_craft",
    "F_operator", "F_elementary", "F_total", "sep1",
    "M_armed", "M_manager", "M_intellectual", "M_technical",
    "M_clerical", "M_sales", "M_agriculture", "M_craft",
    "M_operator", "M_elementary", "M_total", "sep2",
    "T_armed", "T_manager", "T_intellectual", "T_technical",
    "T_clerical", "T_sales", "T_agriculture", "T_craft",
    "T_operator", "T_elementary", "T_total"),
  breakdown_var = "occupation",
  cat_labels  = c(
    armed       = "Armed Forces",
    manager     = "Managers",
    intellectual = "Intellectual professions",
    technical   = "Technical professions",
    clerical    = "Clerks",
    sales       = "Sales & services",
    agriculture = "Agriculture",
    craft       = "Craft & trades",
    operator    = "Machine operators",
    elementary  = "Elementary occupations",
    total       = "total"))
 
# seniority brackets (Tav.5, 10, 16)
# 8 brackets (0-4, 5-9, 10-14, 15-19, 20-24, 25-29, 30-34, 35+) + total
meta_seniority <- list(
  raw_names = c(
    "row_label",
    "F_yr0004", "F_yr0509", "F_yr1014", "F_yr1519",
    "F_yr2024", "F_yr2529", "F_yr3034", "F_yr35plus", "F_total", "sep1",
    "M_yr0004", "M_yr0509", "M_yr1014", "M_yr1519",
    "M_yr2024", "M_yr2529", "M_yr3034", "M_yr35plus", "M_total", "sep2",
    "T_yr0004", "T_yr0509", "T_yr1014", "T_yr1519",
    "T_yr2024", "T_yr2529", "T_yr3034", "T_yr35plus", "T_total"),
  breakdown_var = "tenure",
  cat_labels    = c(
    yr0004 = "0-4 years", yr0509 = "5-9 years",
    yr1014 = "10-14 years", yr1519 = "15-19 years",
    yr2024 = "20-24 years", yr2529 = "25-29 years",
    yr3034 = "30-34 years", yr35plus = "35+ years",
    total = "total"))

# age brackets (table 14) 
# three brackets - 14 to 29, 30 to 49 and 50+
meta_age <- list(
  raw_names = c("row_label",
    "F_young", "F_mid", "F_old", "F_total", "sep1",
    "M_young", "M_mid", "M_old", "M_total", "sep2",
    "T_young", "T_mid", "T_old", "T_total"),
  breakdown_var = "age",
  cat_labels    = c(young = "14-29 years", mid = "30-49 years",
                    old   = "50+ years",   total = "total"))
```

## One generalizing function

-   Input: sheet name, meta information (as in column names + breakdown
    label), outcome label (hourly_wage / annual_wage / annual_hours)

-   Output: tidy long-format tibble

```{r general function}
parse_ses_table <- function(sheet_name, meta, outcome_label) {
  n_target <- length(meta$raw_names)

  raw <- read_excel(
    "data/Tavole_Struttura-delle-retribuzioni_2022.xlsx",
    sheet     = sheet_name,
    col_names = FALSE,
    col_types = "text",
    skip      = 4
  )

  # keep only the columns described by the meta; rename
  raw <- raw[, seq_len(n_target)]
  names(raw) <- meta$raw_names
 
  # detect and propagate section types
  cleaned <- raw |>
    mutate(strat_type = if_else(
      is.na(row_label) & str_trim(.data[[meta$raw_names[2] ]]) %in% sections,
      str_trim(.data[[ meta$raw_names[2] ]]),
        NA_character_)) |>
    fill(strat_type, .direction = "down") |>
    # remove section marker rows, footnotes, and trailing blank rows
    filter(
      !is.na(row_label),
      !str_detect(row_label, fixed("legenda")),
      !str_detect(row_label, fixed("Fonte"))) |>
    # recode section labels
    mutate(strat_type = recode(str_trim(strat_type),
      "ATTIVITÀ ECONOMICA"          = "sector",
      "ATTIVITÀ ECONOMICA "         = "sector",
      "RIPARTIZIONE GEOGRAFICA"     = "geography",
      "CLASSE DIMENSIONALE"         = "firm_size",
      "TIPO DI CONTROLLO ECONOMICO" = "control_type"))
 
  # drop separator columns
  cleaned <- cleaned |> select(-c(sep1, sep2))
 
  # convert values to numeric and ".." to NA
  value_cols <- setdiff(names(cleaned), c("row_label", "strat_type"))
  cleaned <- cleaned |>
    mutate(across(
      all_of(value_cols),
      ~ as.numeric(if_else(str_trim(.) == "..", NA_character_, .))
    ))
 
  # pivot to long 
  pivot_long <- cleaned |>
    pivot_longer(
      cols = all_of(value_cols),
      names_to  = c("sex", "breakdown_cat"),
      names_sep = "_",
      values_to = "value") |>
    mutate(
      sex = recode(sex, "F" = "Female", "M" = "Male", "T" = "Total"),
      # map short breakdown_cat codes to readable labels
      breakdown_cat_label = recode(breakdown_cat, !!!meta$cat_labels),
      breakdown_var = meta$breakdown_var,
      outcome_var  = outcome_label,
      table_id = sheet_name)
 
  pivot_long
}
```

Test it on table 2.

```{r}
tav2_test <- parse_ses_table(
  sheet_name  = "Tav.2",
  meta = meta_contract_type,
  outcome_label = "hourly_wage"
)
```

Since it worked, let's map all tables to their specific attributes that
are needed in the functions. These information are taken from each
table, that specify the type of outcome and breakdown variable they
refer to.

```{r}
table_map <- tribble(
  ~sheet,    ~meta,             ~outcome,
  "Tav.1",   "education",         "hourly_wage",
  "Tav.2",   "contract_type",     "hourly_wage",
  "Tav.3",   "hours_regime",      "hourly_wage",
  "Tav.4",   "occupation",        "hourly_wage",
  "Tav.5",   "tenure",            "hourly_wage",
  "Tav.6",   "education",         "annual_wage",
  "Tav.7",   "contract_type",     "annual_wage",
  "Tav.8",   "hours_regime",      "annual_wage",
  "Tav.9",   "occupation",        "annual_wage",
  "Tav.10",  "tenure",            "annual_wage",
  "Tav.11",  "education",         "annual_hours",
  "Tav.12",  "contract_type",     "annual_hours",
  "Tav.13",  "hours_regime",      "annual_hours",
  "Tav.14",  "age",               "annual_hours",
  "Tav.15",  "occupation",        "annual_hours",
  "Tav.16",  "tenure",            "annual_hours"
)
 
# named list of all meta vars for lookup
meta_list <- list(
  education = meta_education_level,
  contract_type = meta_contract_type,
  hours_regime = meta_hours,
  occupation   = meta_ISCO,
  tenure = meta_seniority,
  age = meta_age
)
```

# Final SES dataset

Let's try to have the final dataset and see if it works:

```{r}
ses_data <- pmap(
  table_map,
  function(sheet, meta, outcome) {
    parse_ses_table(
      sheet_name = sheet,
      meta = meta_list[[meta]],
      outcome_label = outcome
    )
  }
) |>
  bind_rows()

ses_data |> count(table_id, breakdown_var, outcome_var)
```

What these processes and function gave me:

-   table_id; from which ISTAT source table the row came from

-   strat_type; the primary dimension the table stratifies workers on. 4
    values:

    -   sector: NACE Rev. 2 economic activity

    -   firm_size: number-of-employees bands

    -   geography: Italian macro-region

    -   control_type: public vs. private ownership of companies

-   row_label; the specific category within strat_type (53 unique).
    Concretely:

    -   for *sector*: NACE letter/double-letter codes B–S with Italian
        descriptions (e.g. "C - Attività manifatturiere", "K - Attività
        finanziarie e assicurative").

    -   for *geography*: Nord-Ovest, Nord-Est, Centro, Sud, Isole,
        'Totale' (total; maybe to drop?)

    -   for *firm_size*: 10-49, 50-249, 250-499, 500-999, 1000+,
        'Totale' = total (plus two stray Controllo pubblico/privato rows
        that look like leakage from a firm-size × control_type table —
        worth a sanity check).

    -   for control_type: public control (controllo pubblico), private
        control (controllo privato), and total (also maybe to be
        dropped).

-   sex; Female, Male, Total.

-   breakdown_var; the secondary dimension cross-tabulated against
    strat_type. 5 'meta' values:t

    -   *tenure*: years with current employer

    -   *occupation*: ISCO-08 1-digit major group

    -   *education:* ISCED level

    -   *contract_type*: permanent vs. temporary

    -   *hours_regime*: full-time vs. part-time

-   breakdown_cat; short machine-friendly code for the secondary
    category (26 unique). Grouped by breakdown_var:

    -   education: low, med, high

    -   contract_type: permanent, temporary

    -   hours_regime: part-time, fullt-ime

    -   tenure: yr0004, yr0509, yr1014, yr1519, yr2024, yr2529, yr3034,
        yr35plus

    -   occupation (ISCO-08 major groups): armed, manager, intellectual,
        technical, clerical, sales, agriculture, craft, operator,
        elementary, total

-   value; the numeric statistic itself. Always interpret in the unit
    implied by outcome_var. No missing values in the file.

'breakdown_cat_label' was developed as a human-readable version of
'breakdown_cat' (e.g. low --\> ≤ lower secondary, yr35plus --\> 35+
years, manager --\> Managers etc.), and then kept as the 'breakdown_cat'
variable.

```{r}
ses_data <- ses_data |>
  mutate(breakdown_cat = breakdown_cat_label) |>  
  select(-breakdown_cat_label)

# mutate the regional labeling so that it matches the RACLI organization
ses_data <- ses_data |> mutate(
    nuts1 = case_when(
      strat_type == "geography" & str_detect(row_label, "Nord-Ovest|Nord-ovest") ~ "Nord-Ovest",
      strat_type == "geography" & str_detect(row_label, "Nord-Est|Nord-est")     ~ "Nord-Est",
      strat_type == "geography" & str_detect(row_label, "Centro")                ~ "Centro",
      strat_type == "geography" & str_detect(row_label, "Sud")                   ~ "Sud",
      strat_type == "geography" & str_detect(row_label, "Isole")                 ~ "Isole",
      TRUE ~ NA_character_
    ))


# write_csv(ses_data, file = "ses_data.csv")
```

This data set gives me the full sectorial and demographic structure of
Italian wages in 2022.

# 0.1 Regional-level disaggregation

The final goal is to have these data sets:

-   ses_clean = wages by sector / firm size / control type at NUTS-1 (5
    macro geographical areas)

-   racli_clean = wages by region/province × sex × year × quantile at
    NUTS-2 (20 regions, 107 provinces)

-   region_keys = combination of NUTS-1 macro-areas and NUTS-2 regions
    (bridge between the sets)

## Adding RACLI regional data

The RACLI data set has four main dimensions: geography (137 territories:
country / NUTS-1 geographical macro areas / 20 NUTS-2 regions / \~107
provinces), sex (Male / Female / Total), year (2014–2023, that I
filtered 2022) and wage statistics (mean / p10 / median / p90).

Differently from the SES data, it only has information regarding every
employed person in a region, regardless of education level or
occupation.

```{r}
library(readr)
racli_raw <- read_csv("data/Sesso - prov (IT1,533_957_DF_DCSC_RACLI_8,1.0).csv",
                      quote = "")

racli_raw <- racli_raw |>
  filter(TIME_PERIOD == 2022) # filter 2022
```

Cleaning the variables.

```{r}
racli_clean <- racli_raw |>
  select(geo_code  = REF_AREA,        # ITC, ITC1, ITC11 etc.
         geo_label  = Territorio,      # specific geographical references
         indicator = DATA_TYPE,       # HOUWAG_ENTEMP_AV_MI / FIRD / MED 
         sex_code = SEX,             # 1 males, 2 females, 9 total
         year  = TIME_PERIOD,
         value = Osservazione) |>
  mutate(
    # indicator codes turned into short readable names
    indicator = recode(
      indicator,
      "HOUWAG_ENTEMP_AV_MI" = "mean",
      "HOUWAG_ENTEMP_FIRD_MI" = "p10",
      "HOUWAG_ENTEMP_MED_MI"  = "median",
      "HOUWAG_ENTEMP_NIND_MI" = "p90"
    ),
    # sex codes to match labels in ses_clean
    sex = recode(
      as.character(sex_code),
      "1" = "Male",
      "2" = "Female",
      "9" = "Total"
    ),
    # classify each territory by its NUTS code length:
    #   2 chars (IT)       -> country
    #   3 chars (ITC)      -> NUTS-1 macro-area
    #   4 chars (ITC1)     -> NUTS-2 region
    #   5 chars (ITC11)    -> NUTS-3 province
    geo_level = case_when(
      nchar(geo_code) == 2 ~ "country",
      nchar(geo_code) == 3 ~ "nuts1",
      nchar(geo_code) == 4 ~ "nuts2",
      nchar(geo_code) == 5 ~ "nuts3",
      TRUE                 ~ NA_character_),
    # derive the parent NUTS-1 code for every row (truncate to first 3 chars)
    nuts1_code = if_else(nchar(geo_code) >= 3, substr(geo_code, 1, 3), NA_character_)) |>
  select(geo_code, geo_label, geo_level, nuts1_code, sex, year, indicator, value)

racli_clean
```

## Creating a regional code book

In a specific data set, mostly developed as a 'code book' for reference,
each region will be matched to its macro NUTS-1 area, as well as to its
local code, used in the RACLI data set. It will be a reference for the
codes present both in the RACLI and in the SES data frames.

```{r}
# building directly from the RACLI data so the codes are consistent
regional_codes <- racli_clean |>
  filter(geo_level == "nuts2") |>
  distinct(
    nuts2_code  = geo_code,
    nuts2_label = geo_label,
    nuts1_code) |>
  mutate(
    nuts1_label = recode(
      nuts1_code,
      "ITC" = "Nord-Ovest / North-West",
      "ITD" = "Nord-Est / North-East",   # old NUTS code (used by RACLI)
      "ITH" = "Nord-Est / North-East",   # current NUTS code (post-2021 revision)
      "ITE" = "Centro / Center",     # old NUTS code (used by RACLI)
      "ITI" = "Centro / Center",     # current NUTS code (post-2021 revision)
      "ITF" = "Sud / South",
      "ITG" = "Isole / Islands")) |>
  arrange(nuts1_code, nuts2_code)

# write_csv(regional_codes, file = "regional_codes.csv")
# write_csv(racli_clean, file = "racli_clean.csv")
```

# PHASE 1

## Exploratory Data Analysis

The EDA covers both datasets before any missingness is introduced. The
goal is to understand their structure, distributions, and any quality
issues, and to produce a first-pass look at the gender wage gap signal
visible in the raw data.

## RACLI data

```{r eda-racli}
plot_intro(racli_clean)
plot_missing(racli_clean)

# bar plot
plot_bar(racli_clean |> select(geo_level, sex, indicator))

# density by sex
racli_clean |>
  filter(sex %in% c("Male", "Female"), indicator == "mean", 
         geo_level == "nuts2") |> 
  ggplot(aes(x = value, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(x = "Mean hourly wage (EUR/hour)", title = "NUTS-2 wage density by sex")
```

```{r eda-racli-2}
# pivot wide
racli_wide_eda <- racli_clean |>
  filter(sex %in% c("Male", "Female")) |>
  select(geo_code, geo_label, geo_level, nuts1_code, sex, indicator, value) |>
  pivot_wider(names_from = indicator, values_from = value)

# histogram
plot_histogram(
  racli_wide_eda |> filter(geo_level == "nuts2") |> select(mean, p10, median, p90))

racli_wide_eda |>
  filter(geo_level == "nuts2") |>
  ggplot(aes(x = nuts1_code, y = mean, fill = sex)) +
  geom_boxplot() +
  labs(x = "NUTS-1 macro-area", y = "Mean hourly wage")

plot_correlation(
  racli_wide_eda |> filter(geo_level == "nuts2") |> select(mean, p10, median, p90))
```

```{r eda-racli-summary}
racli_clean |>
  filter(sex %in% c("Male", "Female")) |>
  group_by(indicator, sex) |>
  summarise(
    mean_val = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    min_val = min(value, na.rm = TRUE),
    max_val = max(value, na.rm = TRUE))
```

Main findings:

-   The dataset is fully complete, no missing values.

-   `geo_level` has 3 levels, but the most prominent one is NUTS-3 (107
    provinces); sex and indicator perfectly balanced in thirds. Female
    distribution in the density plot shows a clear left shift. The two
    genders are overlapping, but distinct peaks reflecting the gender
    gap.

-   From the histogram the wage quantiles at NUTS-2 (regional) level,
    all four statistics roughly right-skewed; p10 clusters around 8-10
    €/hour, the mean around 13-16, and p90 around 18-22 euros. Boxplot
    by NUTS-1 area show that ITC/ITD (North East and West) have higher
    medians and wider spreads compared to the South and the Islands,
    which have it lower and more compressed.

-   In the correlation heatmap all four quantiles show strong
    correlation with each other (\>0.95). The summary table (indicator x
    sex) indicate that male wages exceed female's across every quantile,
    and the gap seems to be somewhat stable across the distribution.

## SES data

```{r eda-ses-intro}
introduce(ses_data)
plot_intro(ses_data)
plot_missing(ses_data)

# bar plot
plot_bar(ses_data |> select(strat_type, sex, breakdown_var, outcome_var))

# histogram
plot_histogram(ses_data |> select(value))

# boxplot
ses_data |>
  filter(sex %in% c("Female", "Male")) |>
  ggplot(aes(x = outcome_var, y = value, fill = sex)) +
  geom_boxplot() +
  labs(x = "Outcome variable", y = "Value")
```

```{r eda-ses-corr}
ses_data |>
  filter(strat_type == "geography", sex == "Total",
         row_label != "Totale", breakdown_cat != "total") |>
  select(nuts1, breakdown_var, breakdown_cat, outcome_var, value) |>
  pivot_wider(names_from = outcome_var, values_from = value) |>
  select(hourly_wage, annual_wage, annual_hours) |>
  drop_na() |>
  plot_correlation()

ses_data |>
  filter(strat_type == "geography", outcome_var == "hourly_wage",
         sex %in% c("Female", "Male"),
         row_label != "Totale", breakdown_cat != "total") |>
  group_by(sex, breakdown_var) |>
  summarise(
    mean_wage = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    min_val = min(value, na.rm = TRUE),
    max_val = max(value, na.rm = TRUE))
```

-   this data set is much larger and more faceted than the RACLI one,
    and has missing information: NUTS-1 (geographical area, in this case
    would be the country) is NA for non-geography strata rows, but it is
    expected by design. Information regarding armed forces is missing in
    some areas as well.

-   The histogram shows a right-skew of the `value,` because hourly
    wages (\~10–20), annual wages (\~15,000–25,000), and annual hours
    (\~1,400–1,800) are mixed on the same axis. It must be
    differentiated through `outcome_var`. The boxplot of sex x
    outcome_var shows that female earn less than males, for hourly and
    annual wages. The annual hours gap is smaller but still present,
    pointing to both a pay rate and a hours-worked component.

-   The correlation heatmap (hourly/annual wage/annual hours) shows a
    strong relationship between annual wage and annual hours, as could
    be expected. The hourly wage moderately correlated with annual wage,
    while hourly wage nearly uncorrelated with annual hours (possibly
    indicating that full-time hours are compressed regardless of pay).

-   In the summary table (geography rows, hourly wage, sex x
    breakdown_var), occupation shows the highest average wages and the
    widest Female-Male spread. Tenure and education both show clear wage
    gradients, while temporary contract and part-time show the lowest
    values.

# PHASE 2

## Measuring gender wage gap

The first phase of this experimental process is to establish a baseline
gender pay gap calculated from real ISTAT 2022 data. First, the
unadjusted gender pay gap will be calculated. It refers to the total
observed wage disparity per region, calculated from RACLI data. The mean
hourly wages of women and men in region r: 1 - (mean hourly wage for
women / mean hourly wave for men in region r). It will also be
calculated on a country and macro area.

```{r unadj-GPG}
# pivot Male/Female mean hourly wages to columns and compute the gap
ugpg <- function(level) {
  racli_clean |>
    filter(indicator == "mean",
           geo_level == level,
           sex %in% c("Male", "Female")) |>
    select(geo_code, geo_label, nuts1_code, sex, value) |>
    pivot_wider(names_from = sex, values_from = value) |>
    mutate(uGPG = 1 - Female / Male)
}

ugpg_country <- ugpg("country")
ugpg_nuts1 <- ugpg("nuts1") |> arrange(desc(uGPG))
ugpg_nuts2 <- ugpg("nuts2") |> arrange(desc(uGPG))

ugpg_country
ugpg_nuts1
ugpg_nuts2

# SAVE  
# write_csv(ugpg_country, file = "ugpg_country.csv")
# write_csv(ugpg_nuts1, file = "ugpg_nuts1.csv")
# write_csv(ugpg_nuts2, file = "ugpg_nuts2.csv")
```

On the national level, women earn \~9.87% less per hour than men. When
looking at the pay gap in the macro areas, surprisingly the gender pay
gap is higher in the northern areas, where women tend to earn more than
compared to the center and southern areas, but the difference compared
with men is staggering. The bigger difference is found in the North-West
area, where women earn approximately 13% less than men, closely followed
by the North-East. The Islands, Sicily and Sardinia, have the smallest
unadjusted pay gap, reaching almost 5%.

On a regional level, the unadjusted regional GPG confirms the expected
pattern: the widest gaps are concentrated in the wealthy Northern
regions, such as Liguria (15.5%), Lombardia (12.9%), Friuli-Venezia
Giulia (12.6%), Emilia-Romagna (12.3%) and Piemonte (12.2%) top the
ranking. The Southern regions cluster at the bottom, with Calabria
(\~5.2%), Molise (5.2%), Campania (6.8%) and Sicilia (7.5%) showing the
narrowest raw gaps.

A "composition-adjusted" GPG will be calculated next. Using the SES
data, for each NUTS-1 macro area and each breakdown category (education
level, contract type, hours regime, occupation, tenure bracket), compute
the gender gap.

```{r gender-gap-sections}
within_gaps <- ses_data |>
  filter(strat_type == "geography",
         outcome_var == "hourly_wage",
         row_label != "Totale",             # drop the national-total row
         breakdown_cat != "total",    # drop within-breakdown totals
         sex %in% c("Female", "Male")) |>
  select(area = row_label,
         breakdown_var, breakdown_cat,
         sex, value) |>
  pivot_wider(names_from = sex, values_from = value) |>
  filter(!is.na(Female), !is.na(Male), Male > 0) |>
  mutate(gap = 1 - Female / Male)

within_gaps
```

aGPG per NUTS-1 area, per breakdown variable. Weight within-category
gaps by the female-wage share of that category within the area, then sum
across categories.

```{r adj-GPG-area-breakdown}
agpg_area_breakdown <- within_gaps |>
  group_by(area, breakdown_var) |>
  mutate(weight = Female / sum(Female, na.rm = TRUE)) |>
  summarise(aGPG = sum(weight * gap, na.rm = TRUE),
            n_categories = n(),
            .groups = "drop") |>
  arrange(area, breakdown_var)

agpg_area_breakdown

# write_csv(agpg_area_breakdown, file = "agpg_area_breakdown.csv")
```

The composition-adjusted GPG reveals two important patterns. First,
occupation and education are the dimensions where the gap remains
largest and most stable across all macro-areas (occupation: \~10–19%;
education: \~6–18%), confirming that even within the same job category
or education level, men out-earn women. These are the dimensions where
the "unexplained" residual is hardest to wash away. Second, and most
strikingly, the South and the islands flip negative on contract_type,
hours_regime, and tenure (women out-earn men within
full-time/permanent/longer-tenured cells), while these gaps stay
positive in the central regions, North-East and North-West

A single aGPG number per area, taken as the mean across the five
breakdown variables.

```{r adj-GPG-area}
agpg_area <- agpg_area_breakdown |>
  group_by(area) |>
  summarise(aGPG = mean(aGPG), .groups = "drop") |>
  arrange(desc(aGPG))

agpg_area

# write_csv(agpg_area, file = "agpg_area.csv")
```

These tables are the ground-truth baseline against which the imputed
(MICE / LLM) datasets will later be evaluated in Step 4.

!! higher gap in more productive and richer regions –\> SELECTION
EFFECT - to be looked into for theoretical background !!

# 2.1 Regional disadvantages

A regional disadvantage score (reg_dis) is built from the RACLI 'sex ==
total' cells at NUTS-2 micro level, using two indicators:

-   the first is the regional mean hourly wage, where a low value
    corresponds to a disadvantaged labor market; this is therefore
    reverse-signed and z-standardized, so that higher scores mean lower
    wages.

-   the second is the p90/p10 ratio per region, a standard measure of
    wage dispersion in which a high value signals a more polarized labor
    market (a wider gap between top and bottom earners), and is
    z-standardized in its observed direction.

The two standardized measures are then averaged into a single data frame
'reg_dis' per NUTS-2 region, which is finally categorized into three
labels, low / moderate / high, to give discrete disadvantage levels and
clearer indicators. As a result, this score measures how structurally
disadvantaged a region's labor market is, by combining how low its wages
are with how unequally those wages are distributed. This index will be
used to fabricate MNAR missingness among the data, to further test
recovery methods; MICE with PMM and LLM's reconstruction abilities.

```{r regional-disadv-score}
reg_dis <- racli_clean |>
  filter(geo_level == "nuts2",
         sex == "Total",
         indicator %in% c("mean", "p10", "p90")) |>
  select(geo_code, geo_label, nuts1_code, indicator, value) |>
  pivot_wider(names_from = indicator, values_from = value) |>
  mutate(
    p90_p10 = p90 / p10,
    # z-standardize each indicator; low wage and high dispersion = disadvantaged
    z_wage  = -as.numeric(scale(mean)),     # reverse-signed: low wage -> high score
    z_disp  =  as.numeric(scale(p90_p10)),  # "high = bad" direction
    # composite: simple average of the two standardized indicators
    reg_dis = (z_wage + z_disp) / 2,
    # labels
    reg_dis_lvl = cut(reg_dis,
                      breaks = quantile(reg_dis, c(0, 1/3, 2/3, 1)),
                      labels = c("low", "moderate", "high"),
                      include.lowest = TRUE)) |>
  arrange(desc(reg_dis))

reg_dis 

# write_csv(reg_dis, file = "reg_dis.csv")
```

!! Important to note: the 'sex == total' mean wage used here is a
weighted average of male and female wages, so 'reg_dis' is not fully
independent of the unadjusted GPG as the two share underlying data.

# PHASE 3

## MCAR pilot on RACLI

To test the ability and accuracy of LLM's imputation methods, I will
first test the basic MCAR missingness scenario: missing completely at
random.

Under the MCAR assumption, there is no relationship between missingness
and either the wage or the region, as each cell has the same probability
of being missing. This is therefore ideal as a baseline for MAE and
RMSE, it is the baseline 'noise'. It will be calculated both on the
RACLI data and on the

STEPS:

1.  Pivot RACLI data wide so MICE has structural predictors to work
    with. One row per region × sex, with 'mean' , ' p10', 'median',
    'p90' as columns.

2.  A new data set where about 20% of the 'mean' column is missing
    completely at random (MCAR).

3.  Run MICE with predictive mean matching (PMM) as the standard
    practice. PMM here selected because it is robust, and because it
    draws imputed values from observed values rather than from a fitted
    distribution: every imputed wage is a real euro/hour figure within
    the empirical support.

4.  Pivot back to long, recompute the GPG measures using the already
    present function, as a comparison to understand how well the PMM and
    MICE metrics were at reconstructing the missing data.

```{r MICE-imputation-RACLI}
library(mice)
set.seed(240603) # for reproducibility

# RACLI wide format 
racli_wide <- racli_clean |>
  filter(sex %in% c("Male", "Female")) |>
  select(geo_code, geo_label, geo_level, nuts1_code, sex, indicator, value) |>
  pivot_wider(names_from = indicator, values_from = value)

racli_wide  # ground truth, keep aside

# MCAR mask: ~20% of mean cells set to NA, independently of anything
miss_rate <- 0.20
racli_missing <- racli_wide |>
  mutate(mean = if_else(runif(n()) < miss_rate, NA_real_, mean))

# SAVE THE GROUND TRUTH - MNAR missingness will be then applied to it
# write_csv(racli_wide, file = "racli_wide_original.csv")
```

Now about 20% of the 'mean' column has missing data completely at
random.

## MICE imputation on pilot MCAR

Imputation on the RACLI with the engineered missingness completely at
random.

```{r MCAR-mice-RACLI}
# MICE with PMM where predictors: the other quantiles + sex + geographical factors
mice_imputed <- racli_missing |>
  mutate(across(c(sex, geo_level, nuts1_code), as.factor)) |>
  select(sex, geo_level, nuts1_code, mean, p10, median, p90)

mice_fit <- mice(mice_imputed, m = 5, method = "pmm", printFlag = FALSE)

# pool the 5 imputed data sets by averaging the imputed 'mean' column
imp_means <- sapply(1:5, function(i) complete(mice_fit, i)$mean)
racli_imp <- racli_missing |>
  mutate(mean = rowMeans(imp_means))

# SAVE
# write_csv(racli_imp, file = "racli_baseline_mcar.csv")
```

The same process will be performed on the SES dataset.

## MCAR pilot on SES

The missingness completely at random will be engineered as pilot for the
SES data, specifically on the data relevant for the composition-adjusted
GPG. This pilot run will establish the baseline noise of the imputation
pipeline on this dataset.

The steps followed for this step will be similar to the ones executed
when establishing the baseline noise for the RACLI data, but the
missingness here will be around 5% of the data. !! ACCORDING TO THE
LITERATURE !!

```{r}
library(mice)
set.seed(240603) 

# pivoted dataset and filtered to only keep the rows where NUTS1 are specified
ses_wide <- ses_data |>  
  filter(strat_type == "geography",
         outcome_var == "hourly_wage",
         row_label != "Totale",
         breakdown_cat != "total",
         sex %in% c("Female", "Male")) |>
  select(area = row_label,
         breakdown_var, breakdown_cat,
         sex, value) |>
  pivot_wider(names_from = sex, values_from = value) |>
  filter(!is.na(Female), !is.na(Male), Male > 0)

# create the baseline corrupted MCAR dataset
miss_rate_ses <- 0.05
ses_missing <- ses_wide |> 
  mutate(
    Female = if_else(runif(n()) < miss_rate_ses, NA_real_, Female),
    Male   = if_else(runif(n()) < miss_rate_ses, NA_real_, Male)
  )

# SAVE ses_wide as ground truth - MNAR missingness will be then applied to it
# write_csv(ses_wide, file = "ses_wide_original.csv")
```

## MICE imputation on pilot SES

MICE imputation using PMM. The SES dataset was filtered to the 5 sheets
reporting hourly wages (Tav. 1–5). Within each sheet, only the geography
stratification rows were kept with the specified 5 NUTS-1 macro-areas
(Nord-Ovest, Nord-Est, Centro, Sud, Isole), and "Total" dropped. Each of
those rows is crossed with the breakdown categories from each table.

```{r}
# convert descriptors to factors so MICE can utilize them as predictors
mice_ses_input <- ses_missing |>
mutate(across(c(area, breakdown_var, breakdown_cat), as.factor)) |>
select(area, breakdown_var, breakdown_cat, Female, Male)

# run the imputation over 5 iterations
mice_ses_fit <- mice(mice_ses_input, m = 5, method = "pmm", printFlag =
FALSE)

# pooling the 5 imputed datasets
imp_female_matrix <- sapply(1:5, function(i) complete(mice_ses_fit,
i)$Female)
imp_male_matrix <- sapply(1:5, function(i) complete(mice_ses_fit,
i)$Male)

ses_imputed <- ses_missing |>
mutate(
Female = rowMeans(imp_female_matrix),
Male = rowMeans(imp_male_matrix)
)

```

Now I have the imputed pilot datasets to establish algorithmic noise for
both the RACLI data (from which the unadjusted gpg was calculated) SES
data (from which the "composition-adjusted" gpg was measured).

## Recreating and comparing GPG metrics

First, let's look at how the uadjusted GPG was recreated and let's
evaluate it against the original metric. Here bias is computed per
region as 'uGPG_imp' − 'uGPG_true', where 'uGPG_true' is the baseline
gap calculated from the complete RACLI data.

```{r}
# recompute uGPG on imputed data
ugpg_from_imp <- function(level) {
  racli_imp |>
    filter(geo_level == level) |>
    select(geo_code, geo_label, nuts1_code, sex, mean) |>
    pivot_wider(names_from = sex, values_from = mean) |>
    mutate(uGPG = 1 - Female / Male)
}

ugpg_nuts2_imp <- ugpg_from_imp("nuts2")

compare_nuts2 <- ugpg_nuts2 |>
  select(geo_code, geo_label, uGPG_true = uGPG) |>
  inner_join(ugpg_nuts2_imp |> select(geo_code, uGPG_imp = uGPG),
             by = "geo_code") |>
  mutate(bias      = uGPG_imp - uGPG_true,
         abs_error = abs(bias))

summary_metrics <- compare_nuts2 |>
  summarise(
    mean_bias = mean(bias),
    MAE       = mean(abs_error),
    RMSE      = sqrt(mean(bias^2)),
    spearman  = cor(uGPG_true, uGPG_imp, method = "spearman")
  )

summary_metrics
compare_nuts2 |> arrange(desc(abs_error))
```

This initial baseline attempt behaves very well. Mean bias across the 20
micro-level regions is essentially zero (0.0005), MAE is 0.3 percentage
points, RMSE equals to 0.007, and the Spearman rank correlation between
true and imputed regional uGPGs is 0.96, indicating that regional
rankings are almost perfectly preserved.

The fact that the substantive ordering of regions survives the MCAR
missing data method is especially reassuring, since the focus of the GPG
depends more on which regions stand out than on the exact decimal of
each gap. Taken together, these results establish the noise floor of the
pipeline: the residual error attributable to the imputation procedure
itself, before any informative missingness is introduced. The most
relevant takeaway of this initial MCAR test on PMM is that any possible
deterioration observed in the following steps of experiment can be
attributed to the missingness mechanism rather than to limitations of
the MICE-PMM method as such.

-   well performing baseline when the NAs are missing completely at
    random and retreived through MICE imputation (PMM).

Now, from the imputed data, the composition-adjusted GPG will be
calculated and compared to the one obtained based on the real SES data.

```{r}
# compute the within-category gaps and weight them by the female-wage share
agpg_breakdown_imp <- ses_imputed |>
  group_by(area, breakdown_var) |>
  mutate(
    gap    = 1 - Female / Male,
    weight = Female / sum(Female, na.rm = TRUE)) |>
  summarise(
    aGPG_imp     = sum(weight * gap, na.rm = TRUE),
    n_categories = n(),
    .groups      = "drop")

# aggregate to a single aGPG value per NUTS-1 macro-area (mean across 5 variables)
agpg_area_imp <- agpg_breakdown_imp |>
  group_by(area) |>
  summarise(aGPG_imp = mean(aGPG_imp), .groups = "drop")

# evaluation at the Area × Breakdown level (25 data points for robust metrics)
compare_ses_breakdown <- agpg_area_breakdown |>
  select(area, breakdown_var, aGPG_true = aGPG) |>
  inner_join(agpg_breakdown_imp |> select(area, breakdown_var, aGPG_imp),
             by = c("area", "breakdown_var")) |>
  mutate(bias      = aGPG_imp - aGPG_true,
         abs_error = abs(bias))

summary_ses_breakdown_metrics <- compare_ses_breakdown |>
  summarise(mean_bias = mean(bias),
            MAE = mean(abs_error),
            RMSE = sqrt(mean(bias^2)),
            spearman = cor(aGPG_true, aGPG_imp, method = "spearman"))

# evaluation at the macro Area level (5 NUTS-1 regions)
compare_ses_area <- agpg_area |>
  select(area, aGPG_true = aGPG) |>
  inner_join(agpg_area_imp |> select(area, aGPG_imp), by = "area") |>
  mutate(bias = aGPG_imp - aGPG_true,
         abs_error = abs(bias))
```

This initial baseline attempt also behaves really well. Mean bias across
the 25 micro-level area-breakdown strata is essentially zero (-0.0002),
MAE is under 0.1 percentage points (0.00096), RMSE equals 0.0014, and
the Spearman rank correlation between true and imputed stratified aGPGs
is 0.999, indicating that the composition-adjusted rankings are almost
perfectly preserved. Looking at the macro-area aggregates, absolute
errors remain low, peaking at just 0.14 percentage points in the *Isole*
macro-area and dropping to near-zero for both the *Sud* and *Nord-est*
zones.

These results establish the noise floor of the SES pipeline, meaning the
residual error attributable to the imputation procedure itself, before
any widespread missingness is introduced. The most relevant takeaway of
this initial MCAR test on PMM is that any future performance changes
observed in the next steps of the experiment can be confidently
attributed to the missingness mechanism, rather than to baseline
limitations of the MICE-PMM performance on this data structure.

-   well performing baseline when the NAs are missing completely at
    random and retreived through MICE imputation (PMM), for SES data as
    well.

The next step is to create the datasets (one for SES and one for RACLI)
for MICE and the LLM to actually reproduce. In the case on RACLI, the
dataset will be objected to MNAR (missing not at random), while for the
Structure of Earnings data, the dataframe will be subjected to a higher
amount of data missing completely at random, MCAR.

# 3.1 Engineered MNAR dataset (RACLI)

Using the regional disaggregation metric that was developed for each
region, the information will be modified in order to fabricate the
missing not at random scenario. This incomplete data will be imputed
both using the standard MICE package, and using the LLMs.

MNAR is the method selected to define missingness because we are
assuming that the probability of the data cells being missing are
directly related to the lack of information/difficulty of collecting
data/increased inequalities in the regions with the lowest regions +
lowest female participation. It is a structural issue rather than a
random one, hence the construction of the regional disadvantage levels.
Here, this will be fabricated by dropping data from the *'racli_clean'*
data (wide format), where the 'mean' cells will be dropped with a
probability scaled according to the regional disadvantage measured.

-   low reg_dis level = \~0.10 probability of mean values being missing

-   medium reg_dis level = \~0.25 probability of mean values being
    missing

-   high reg_dis level = \~0.45 probability of mean values being missing

!! How do I justify/specify this selection of probability scores?

```{r MNAR-reg_dis}
set.seed(240603)

# average drop rate across tiers ~25%, comparable to the MCAR 20% baseline
p_miss_mnar <- c(low = 0.10, moderate = 0.25, high = 0.45)

# tier lookup defined at NUTS-2; NUTS-3 provinces will inherit the
# disadvantage tier of their NUTS-2 region, so missingness propagates 
# down the geographic hierarchy
tier_lookup <- reg_dis |>
  select(geo_code, reg_dis_lvl)

# geo code matching
racli_wide_mnar <- racli_wide |>
  mutate(
    # each row's NUTS-2 region: itself if NUTS-2, truncated 
    # geo_code if NUTS-3, NA otherwise (NUTS-1 / country stay untouched)
    nuts2_parent = case_when(geo_level == "nuts2" ~ geo_code,
                             geo_level == "nuts3" ~ substr(geo_code, 1, 4),
                             TRUE ~ NA_character_)) |>
  left_join(tier_lookup, by = c("nuts2_parent" = "geo_code"))

# Bernoulli draw conditional on tier; aggregate / country rows never masked
racli_exp_mnar <- racli_wide_mnar |>
  mutate(p_drop = p_miss_mnar[as.character(reg_dis_lvl)],
         p_drop = if_else(is.na(p_drop), 0, p_drop),
         drop = runif(n()) < p_drop,
         mean = if_else(drop, NA_real_, mean)) |>
  select(-nuts2_parent, -p_drop, -drop)  # tidy up

plot_missing(racli_exp_mnar)

# SAVE
# write_csv(racli_exp_mnar, file = "racli_exp_mnar.csv")
```

Each cell in the `mean` column is dropped with a probability that
depends on its region's disadvantage tier: roughly 10% in `low`, 25% in
`moderate`, and 45% in `high`, so that the presence of NAs is
concentrated in structurally weaker labor markets, but never
deterministic. NUTS-3 provinces inherit the tier of their NUTS-2 region,
while NUTS-1 and country rows are left intact. `racli_missing_mnar` is
the dataset that MICE and the LLM will both later attempt to
reconstruct.

# 3.2 Engineered MNAR dataset (SES)

For the SES data, the missingness will be fabricated through MNAR at the
macro-area level according to the (adapted) regional disadvantage index.
Only the rows with the specified stratification type at the geographic
level will be kept (already specified with ses_wide).

```{r MCAR-ses}
set.seed(2026) # seed for reproducibility

p_miss_step <- 0.20

ses_exp_mnar <- ses_wide |>
  rename(nuts1 = area) |>          # `area` has the NUTS-1 macro-area label
  filter(!is.na(nuts1)) |>         # keep ONLY rows with a defined NUTS-1 area
  mutate(
    # independent cell-level draws for Female / Male missingness
    drop_F = runif(n()) < p_miss_step,
    drop_M = runif(n()) < p_miss_step,
    Female = if_else(drop_F, NA_real_, Female),
    Male  = if_else(drop_M, NA_real_, Male)) |>
  select(-drop_F, -drop_M)         # tidy up

plot_missing(ses_exp_mnar)

# SAVE EXPERIMENTAL DATASET
# write_csv(ses_exp_mnar, file = "ses_exp_mnar.csv")
```

Now the two experimental data sets, where missingness was engineered,
are created and can be used to test the two retrieval methods: MICE and
LLM.

# PHASE 4

# Baseline LLM & Uniform Function

Before actually starting the experiment, a baseline analysis of the LLM
will be carried out. Based on previous research, the selected LLM will
be OLMo 3: "a competitive, truly Open Language Model, to enable the
scientific study of language models. Unlike most prior efforts that have
only released model weights and inference code, we release OLMo
alongside open training data and training and evaluation code." It also
has a lower verbatim memorization rate, making it the least likely to
possess pre-existing knowledge of specific records that would bias the
results.

<https://arxiv.org/html/2402.00838> <https://arxiv.org/html/2512.13961>

OLMo 3 is a neural network with billions of numbers (weights) in it. To
actually run it and use it, the weights have to be loaded into the
memory of a machine with a powerful GPU. For computational constraints,
an API key was obtained for 5 dollars (of tokens) from HuggingFace.com -
a company that maintains a huge open-source community of the same name
that builds tools, machine learning models and platforms for working
with artificial intelligence, with a focus on data science, machine
learning and natural language processing (NLP). Hugging Face is notable
for its NLP transformers library and a platform that allows users to
share models and data sets (from
<https://www.ibm.com/think/topics/hugging-face>).

It will be used through the 'olmo_chat' function.

```{r}
library(httr)
library(jsonlite)

olmo_chat <- function(prompt,
                      model = "allenai/Olmo-3-7B-Instruct",
                      system = NULL,
                      temperature = 0) {

  msgs <- list()
  if (!is.null(system)) msgs <- c(msgs, list(list(role = "system", content = system)))
  msgs <- c(msgs, list(list(role = "user", content = prompt)))

  resp <- POST(
    url = "https://router.huggingface.co/v1/chat/completions",
    add_headers(
      Authorization = paste("Bearer", Sys.getenv("HF_TOKEN")),
      `Content-Type` = "application/json"),
    body = toJSON(list(
      model  = model,
      messages = msgs,
      temperature = temperature), auto_unbox = TRUE),
    encode = "raw")

  stop_for_status(resp)
  content(resp, as = "parsed")$choices[[1]]$message$content
}
```

Sanity check:

```{r}

olmo_chat("What was the approximate average gross hourly wage in Italy in 2022, in euros? 
          Reply with a single number.",
  system = "You are a careful estimator. Reply with only a number, no text.")
```

# 4.1 Evaluation function

An evaluation function was created, as a one single function to apply to
every imputed dataset, and create a uniform and cleaner workflow. It
returns a one-row tibble, the results of which will be added in the
metrics tracker.

```{r evaluate-function}

evaluate_imputation <- function(imp_dataset, label, metric = c("ugpg", "agpg")) {
  metric <- match.arg(metric)

  if (metric == "ugpg") {
    # RACLI: unadjusted gap from mean hourly wage at NUTS-2
    imp <- imp_dataset |>
      filter(geo_level == "nuts2", sex %in% c("Male", "Female")) |>
      select(geo_code, sex, mean) |>
      pivot_wider(names_from = sex, values_from = mean) |>
      mutate(value_imp = 1 - Female / Male) |>
      select(unit = geo_code, value_imp)

    truth <- ugpg_nuts2 |> select(unit = geo_code, value_true = uGPG)

  } else {
    # SES: composition-adjusted gap, within-category gaps weighted by female-wage share
    imp <- imp_dataset |>
      group_by(area, breakdown_var) |>
      mutate(gap = 1 - Female / Male,
             weight = Female / sum(Female, na.rm = TRUE)) |>
      summarise(value_imp = sum(weight * gap, na.rm = TRUE), .groups = "drop") |>
      unite("unit", area, breakdown_var, sep = " | ")

    truth <- agpg_area_breakdown |>
      select(area, breakdown_var, value_true = aGPG) |>
      unite("unit", area, breakdown_var, sep = " | ")
  }

  truth |>
    inner_join(imp, by = "unit") |>
    mutate(bias = value_imp - value_true, abs_error = abs(bias)) |>
    summarise(
      method = label,
      metric = metric,
      n = sum(!is.na(bias)),
      mean_bias = mean(bias, na.rm = TRUE),
      MAE = mean(abs_error, na.rm = TRUE),
      RMSE = sqrt(mean(bias^2, na.rm = TRUE)),
      spearman = cor(value_true, value_imp, method = "spearman", use = "complete.obs"))
}

# metrics tracker
metrics_tracker <- tibble(
  method = character(),
  metric = character(),
  n = integer(),
  mean_bias = double(),
  MAE = double(),
  RMSE = double(),
  spearman = double())
```

# 4.2 LLM Uninformed baseline (RACLI)

OLMo 3 is asked about each missing RACLI cells by region name and sex
only, with zero data context. This defined the prior knowledge that
model carries about Italian regional wages from training, and the
metrics can therefore be used as a baseline. When using the LLM, the
MNAR engineered missingness will be used, instead of using the MCAR as
it was done to establish the baseline MICE performance and metrics. This
is done because the uninformed LLM ignores all observed data, so the
missingness mechanism doesn't change its guesses or mechanism.

In this chunk, MNAR-deleted RACLI cells will be replaced by OLMo's
uninformed guesses: for each missing NUTS-2 region x sex wage, it builds
a prompt asking the model for that 2022 mean hourly wage with no data
given, which is then parsed into a single number from the reply.
Furthermore, the function discards implausible values (outside €6–25) as
NA, producing one estimated wage per gap.

```{r llm-baseline}
library(purrr)

missing_nuts2 <- racli_exp_mnar |>
  filter(is.na(mean), geo_level == "nuts2") |>
  select(geo_code, geo_label, sex)

llm_baseline <- missing_nuts2 |>
  mutate(prompt = paste0(
    "What was the mean gross hourly wage for ", sex,
    " workers in the Italian region of ", geo_label,
    " in 2022? Reply with a single number in EUR per hour, no explanation."),
    mean_imp = map_dbl(prompt, \(p) {
      Sys.sleep(1)
      resp <- olmo_chat(p,
                        system = "You are a careful economic estimator. Reply with only a single number.")
     
      nums <- str_extract_all(resp, "[0-9]+(?:[.,][0-9]+)?")[[1]]
      nums <- nums[!str_detect(nums, "^(19|20)[0-9]{2}$")] # drop 4-digit years so they aren't read as wages
      val  <- as.numeric(str_replace(nums[1], ",", "."))  # NA only if nothing left
    }))

llm_baseline |> select(geo_label, sex, mean_imp)
```

Result: 16 cells that were removed my the MNAR process are here
recovered, and are the baseline knowledge of the LLM. They are to be
joined with the rest of the data and compared with the real information
from ISTAT, and added to the metrics table to keep track of every step.

```{r llm-v0-fill}
racli_llm_baseline <- racli_exp_mnar |>
  left_join(llm_baseline |> select(geo_code, sex, mean_imp),
            by = c("geo_code", "sex")) |>
  mutate(mean = if_else(is.na(mean), mean_imp, mean)) |>
  select(-mean_imp)

# SAVE
# write_csv(racli_llm_baseline, file = "racli_llm_baseline.csv")

metrics_tracker <- bind_rows(
  metrics_tracker,
  evaluate_imputation(racli_llm_baseline, "LLM_uninformed", metric = "ugpg"))

metrics_tracker
```

The NAs that are still present here are at the province level --\> keep
it like this or expand the prompt to recover NUTS-3 level?

The main findings here, compared to the initial MICE with MCAR
imputation, are:

-   the LLM guesses nearly the same hourly wage for men and women, so
    the imputed gender gap collapses toward zero, explaining the −0.20
    bias, the model systematically flattens the gap

-   because only the missing cells are replaced (and the real values are
    added to the LLM imputed), there are strongly differing cases. For
    example, the €8.50 male estimate compared to a real €13.59 female
    value produces a wildly wrong (even negative) uGPG for that region.
    These cases combined explain why RMSE (0.408) is almost double than
    the MAE (0.261), but the error is concentrated in a few blown-up
    regions that differ strongly with the estimated wages by the LLM.

This poses an interesting finding in itself, as the MICE retrieval
baseline performed extremely well. PMM recovered the regional gaps
almost perfectly because the imputed mean cell is tightly correlated
with the observed quantiles in the same row, allowing the method to
establish values through interpolation. The test will be whether this
accuracy survives the structured, disadvantage-driven gaps of the MNAR
scenario.

In contrast, the uninformed LLM had no structure to draw on, and
collapsed the gender gap toward zero, showcasing a weak performance from
the low-memorization model.

# 4.3 LLM Uninformed baseline SES

The same process will be adapted to the missing SES cell (Female or Male
hourly wage in a given macro-area × breakdown category) with no data
context. Because `ses_exp_mnar` has separate `Female` and `Male` columns
that both have NAs, the missing cells are first pivoted to long format
to build one prompt per gap, then the imputed values are pivoted back to
wide before evaluation.

```{r llm-baseline-ses}
ses_exp_mnar2 <- ses_exp_mnar |>
  pivot_longer(cols = c(Female, Male), names_to = "sex", values_to = "value") |>
  filter(is.na(value)) |>
  select(nuts1, breakdown_var, breakdown_cat, sex)

llm_baseline_ses <- ses_exp_mnar2 |>
  mutate(
    prompt = paste0(
      "What was the mean gross hourly wage for ", sex,
      " workers in ", nuts1, ", Italy, in 2022,",
      " in the ", breakdown_var, " category '", breakdown_cat, "'?",
      " Reply with a single number in EUR per hour, no explanation."
    ),
    value_imp = map_dbl(prompt, \(p) {
      Sys.sleep(1)
      resp <- olmo_chat(p,
                        system = "You are a careful economic estimator. Reply with only a single number.")
      
      nums <- str_extract_all(resp, "[0-9]+(?:[.,][0-9]+)?")[[1]]
      nums <- nums[!str_detect(nums, "^(19|20)[0-9]{2}$")]
      as.numeric(str_replace(nums[1], ",", "."))
    }),
    value_imp = if_else(value_imp < 6 | value_imp > 40, NA_real_, value_imp))

llm_baseline_ses |> select(nuts1, breakdown_var, breakdown_cat, sex, value_imp)
```

The imputed values show almost no separation by sex, as for the RACLI
imputation: OLMo placed both female and male wages around the same €8-9
floor (mean €9.31 for women, €9.51 for men), generalizing the female
estimates especially: 21 of 23 collapsed onto the single value €8.12.
The male guesses were only roughly more varied. The model essentially
assigns to men and women the same value, regardless of region or
education level, and the within-cell gender gap is here recorded to be
near zero.

Now it will be merged:

```{r llm-ses}
ses_llm_baseline <- ses_exp_mnar |>
  pivot_longer(cols = c(Female, Male), names_to = "sex", values_to = "value") |>
  left_join(
    llm_baseline_ses |> select(nuts1, breakdown_var, breakdown_cat, sex, value_imp),
    by = c("nuts1", "breakdown_var", "breakdown_cat", "sex")
  ) |>
  mutate(value = if_else(is.na(value), value_imp, value)) |>
  select(-value_imp) |>
  pivot_wider(names_from = sex, values_from = value) |>
  rename(area = nuts1)

# SAVE
# write_csv(ses_llm_baseline, file = "ses_llm_baseline.csv")

metrics_tracker <- bind_rows(
  metrics_tracker,
  evaluate_imputation(ses_llm_baseline, "LLM_uninformed", metric = "agpg"))

metrics_tracker
```

The weak performance highlighted with the RACLI data is here confirmed.
